\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-22 pldi/draft.tex
Pass on Section 5

\section{Implementation}
\label{impl}

This section describes challenges related to the implementation of
\sys. It starts by describing how the \sys front-end translates a
shell script to the dataflow graph model defined in \S\ref{ir}, by
applying transformation on parallelizable regions~\sx{front-end}.  It
closes with the \sys runtime\todo{/backend} and the technical
challenges that were addressed to achieve performance
benefits~\sx{optimizer}.

@@

As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
translates it to a dataflow graph.  However, there are several
components in shell programs that cannot be arbitrarily parallelized
without affecting the program behaviour---\eg commands connected with
a \ttt{\&\&} operator~\sx{bg}.
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.
To address this issue, we introduce the notion of parallelizable
program regions and a translation pass for converting them to dataflow
dataflow graphs.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently and barriers that are natural points for
enforcing synchronization. Consider the following example:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep "foo" > f3 && sort f3 
\end{lstlisting}
% # Replace sort with sth better

@@

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
Conversely, all other operators, such as the \ttt{;} sequential
composition operator, the \ttt{\&\&} logical operator represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

\heading{Translation Pass}

%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}

The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions from the bottom up, translating their
independent components to dataflow graph nodes until a barrier
construct is reached. All subtrees that are not translated to dataflow
graphs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
dataflow graphs and a call to \sys's runtime.

@@

An important consideration is that due to the highly dynamic nature of
the shell, there is information (such as the values of environment
variables) that are not known to \sys at the time of translation.
%% to the the safety properties that are mentioned above cannot be
%% soundly checked statically.
For safety purposes, \sys takes a conservative approach where it does
not parallelize nodes without complete information.  For example, the
translation pass cannot infer that an environment variable that is
given as an argument to a command is not a flag that changes its
parallelizability class,
%% two un-expand strings do not refer to the same file identifier,
and as a result does not parallelize this sub-expression.

\kk{Can we say anything more about the safety considerations? That (i)
  the translation pass ``checks'' that no nodes write/read or
  write/write to the same file (is that necessary?), that (ii) a
  command should not read from a read-once file. (is that necessary?)}

@@

After translating the parallelizable regions of the input script to
dataflow graphs and optimizing these graphs using the transformations
described in \Cref{ir:transformations}, \sys implements the graph by
translating it back to a shell script. In this section we describe how
\sys achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}

The shell is uniquely lazy, in the sense that most commands and pipes
consume their inputs only when they are ready to process more. This
allows the underlying system to be able to support concurrently
executing many more commands than the processing threads of the
underlying system, by scheduling them concurrently, allowing them to
have progress one at a time. This design decision however, severely
restricts parallelism since nodes are often blocked when their
consumers are processing their input. Consider as an example the
following script:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  mkfifo t1 t2
  grep "foo" > t1 & grep "foo" > t2 &
  cat t1 t2
\end{lstlisting}

The \ttt{cat} command will only try to consume input from \ttt{t2}
when it is done reading from \ttt{t1}. This means that the second
\ttt{grep} will block without processing any input until the first
\ttt{grep} is done processing. Instead of addressing this issue for
specific commands, \sys inserts and instants several eager \ttt{relay}
nodes in critical points of the pipeline as seen in
\Cref{fig:eager}. These nodes force their producers to produce output
and not block, while also preserving the existing task-based
parallelism---that would not be preserved say with the \todo{third}
solution in \Cref{fig:eager}.

@@

\heading{Split node implementation}

As described in \Cref{ir:transformations}, \sys transformations
inserts \ttt{split} nodes to expose parallelism in cases where a
parallelizable node only has one input. The \sys \todo{backend} offers
two options for implementing this split node, to satisfy the inherent
tension between task-based parallelism and knowledge of the number of
lines in the \ttt{split} input. In the general case, the number of
lines of a specific edge is not known a-priori (due to commands that
remove or add new lines), and so \sys provides an automatic split that
first counts the number of lines, and then splits its input to the
desired number of outputs. If however, the user knows the number of
lines at the split insertion edge, they can provide it to \sys, that
then uses it to \ttt{split} without having to block and read its whole
input first. \sys also inserts eager relay nodes after all but the
last output of split in order to address the laziness issue that was
described before.

@@

\TODO{Talk about eagerness in split and how we needed to add eager
  nodes to n-1 outputs.}

\heading{Dangling FIFOs and Zombie Producers}

%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}

% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a producer exits after it has produced and
sent all of its output to a pipe\todo{/its output channel}.
% \ttt{SIGPIPE} 
However, if its consumer exits early, the producer needs to be
notified so that it stops \todo{appending to the stream/trying to push
  data to its output channel}.
% ---so that it avoids unnecessary work and buffering challenges in the kernel.
In \unix, this is achieved by an out-of-band error mechanism:
  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
}
primarily because pipes and pipelines are at the heart of its philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:\kk{Proposed alternative: ``Consider the following shell script:''}
\begin{lstlisting}[language=sh, numbers=none]
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

To solve this problem, \sys emits cleanup logic that operates from the
end of the pipeline and towards its start.  The emitted code first
gathers the IDs of the output processes and passes them as parameters
to \ttt{wait}; this causes \ttt{wait} to block only on the output
producers of the dataflow graph.  Right after \ttt{wait}, \sys inserts
a routine that delivers \ttt{PIPE} signals to any remaining processes
upstream.  This epilogue is repeated for every \todo{merge/join} point
in the transformed pipeline \kk{I don't understand the final sentence}.
% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"

\kk{We have to remember to say that for this to work, all nodes in the
  graph must always open their outputs! If not, then this technique
  doesn't work, and we have to monitor if any node dies to do
  something with it. At the moment we ensure that by doing that in our
  nodes, and by assuming that all nodes that we handle do that.}

@@

\kk{Maybe say something about some of the pure fold functions that we
  implemented. Interesting examples include sort and wc.}

>>>>

\section{From scripts to graphs and back again}
\label{impl}

\kk{Less fun section title: Implementation}

This section describes challenges related to the implementation of
\sys. It starts by describing how the \sys front-end translates a
shell script to the dataflow graph model defined in \S\ref{ir} by
identifying and transforming parallelizable regions~\sx{front-end}. It
closes with the \sys runtime\todo{/backend} and the technical
challenges that were addressed to achieve performance
benefits~\sx{optimizer}.

@@

As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
translates it to a dataflow graph.  However, there are several
components in shell programs that cannot be arbitrarily parallelized
without affecting the program behaviour---\eg commands connected with
a \ttt{\&\&} operator~\sx{bg}.
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.
To address this issue, we introduce the notion of parallelizable
program regions and a translation pass for converting them to dataflow
graphs.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently, and barriers that are natural points for
enforcing synchronization. Consider the following example:

\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep "foo" > f3 && sort f3 
\end{lstlisting}
% # Replace sort with sth better

@@

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
Conversely, all other operators, such as the \ttt{;} sequential
composition operator and the \ttt{\&\&} logical operator, represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

\heading{Translation Pass}
%
%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}
%
The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions from the bottom up, translating their
independent components to dataflow graph nodes until a barrier
construct is reached. All subtrees that are not translated to dataflow
graphs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
dataflow graphs and a call to \sys's runtime.

@@

An important consideration is that due to the highly dynamic nature of
the shell, there is information (such as the values of environment
variables) that are not known to \sys at the time of translation.
%% to the the safety properties that are mentioned above cannot be
%% soundly checked statically.
For safety purposes, \sys takes a conservative approach where it does
not parallelize nodes without complete information.  For example, the
translation pass cannot infer that an environment variable that is
given as an argument to a command is not a flag that changes its
parallelizability class,
%% two un-expand strings do not refer to the same file identifier,
and as a result does not parallelize such sub-expressions.

%% \kk{Can we say anything more about the safety considerations? That (i)
%%   the translation pass ``checks'' that no nodes write/read or
%%   write/write to the same file (is that necessary?), that (ii) a
%%   command should not read from a read-once file. (is that necessary?)}

@@

After translating the parallelizable regions of the input script to
dataflow graphs and optimizing these graphs using the transformations
described in \Cref{ir:transformations}, \sys implements each graph by
translating it back to a shell script. Nodes of the graph are
instantiated with their respective commands, and edges are
instantiated as named pipes. In this section, we describe several
technical challenges that \sys addresses to efficiently perform this
transformation, as well as the implementations of the custom \sys
nodes that are added by the auxiliary transformations.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}
%
%
The shell is uniquely lazy, in the sense that most commands and pipes
consume their inputs only when they are ready to process more. This
allows the underlying system to concurrently execute more commands
than the available processing threads,
%% support concurrently executing many more commands than the
%% processing threads of the underlying system,
%% by scheduling them concurrently, allowing them to
%% have progress one at a time.
by interleaving their executions. This design decision however,
severely restricts parallelism since nodes are often blocked when
their consumers are not requesting any input. Consider as an example
the following script:

\begin{lstlisting}[language=sh, numbers=none, escapeinside={($}{$)}]
  mkfifo t1 t2
  grep "foo" > t1 & grep "foo" > t2 &
  cat t1 t2
\end{lstlisting}

\noindent
The \ttt{cat} command will only try to consume input from \ttt{t2}
when it is done reading from \ttt{t1}. This means that the second
\ttt{grep} will block without processing any input until the first
\ttt{grep} is done processing. Instead of addressing this issue for
each command separately, \sys inserts and instantiates eager
\ttt{relay} nodes in critical points of the pipeline as seen in
\Cref{fig:eager}. These nodes eagerly consume their input, forcing
their upstream nodes to produce output when possible,
%% force their producers to
%% produce output and not block,
while also preserving the existing task-based parallelism---that would
not be preserved say with the \todo{third} solution in
\Cref{fig:eager}.

@@

\heading{Split node implementation}
%
As described in \Cref{ir:transformations}, \sys's optimizer inserts
\ttt{split} nodes to expose parallelism in cases where a
parallelizable node only has one input. In order for the split to be
effective in exposing parallelism, it needs to split its input equally
in its outputs, but the size of its input can not be known
a-priori. To satisfy this inherent tension, \sys supports two split
implementations, a general one that can be used if the script contains
commands that could add or remove lines, and an optimized one that can
be used if the size of the split input is known before the script
executes.
%% The \sys \todo{backend} support two split node implementations, to
%% satisfy the inherent tension between the split efficiency and
%% task-based parallelism.
In the general case, since the number of lines of a specific edge is
not known, \sys provides an automatic split that first consumes its
complete input, counts its lines, and then splits it equally to the
desired number of outputs. If however, the user knows the number of
lines at the split insertion edge, they can configure \sys to split
without having to read its whole input, therefore exploiting available
task-based parallelism. Finally, \sys also inserts eager relay nodes
after all but the last output of split in order to address the
laziness issue that was described before.

%% \TODO{Talk about eagerness in split and how we needed to add eager
%%   nodes to n-1 outputs.}

@@

\heading{Dangling FIFOs and Zombie Producers}
%
%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}
%
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a command exits after it has produced and
sent all of its output to
%% a pipe
its output channel.
% \ttt{SIGPIPE} 
However, if its output is a pipe and the pipe reader exits early, the
command needs to be notified so that it stops writing to it.
%% \todo{appending to the stream/trying to push data to its output
%%   channel}.
% ---so that it avoids unnecessary work and buffering challenges in the kernel.
In \unix, this is achieved by an out-of-band error mechanism: the
operating system will deliver a \ttt{PIPE} signal to the producer,
notifying it that the pipe's consumer has exited.  This is different
from errors for other system calls
%% return an error
and unusual compared to non-\unix systems\footnote{For example,
  Windows indicate errors for \ttt{WriteFile} using its return
  code---similar to \ttt{DeleteFile} and other Win32 functions.}
primarily because pipes and pipelines are at the heart of its
philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:
%% \kk{Proposed alternative: ``Consider the following shell script:''}
\begin{lstlisting}[language=sh, numbers=none]
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat}
to exit before opening \ttt{fifo2}.  As a result, the second \ttt{cat}
never receives a \ttt{PIPE} signal that its consumer exited---after
all, \ttt{fifo2} never even had a consumer.  This, in turn, leaves the
second \ttt{cat} unable to make progress, both (i) blocked unable to
write output, and (ii) unaware that its consumer has exited.  Coupled
with \ttt{wait} at the end, the entire snippet reaches a deadlock.

To solve this problem, \sys emits cleanup logic that operates from the
end of the pipeline and towards its start.  The emitted code first
gathers the IDs of the output processes and passes them as parameters
to \ttt{wait}; this causes \ttt{wait} to block only on the output
producers of the dataflow graph.  Right after \ttt{wait}, \sys inserts
a routine that delivers \ttt{PIPE} signals to any remaining processes
upstream.
%% This epilogue is repeated for every \todo{merge/join} point
%% in the transformed pipeline \kk{I don't understand the final sentence}.
% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"

%% \kk{We have to remember to say that for this to work, all nodes in the
%%   graph must always open their outputs! If not, then this technique
%%   doesn't work, and we have to monitor if any node dies to do
%%   something with it. At the moment we ensure that by doing that in our
%%   nodes, and by assuming that all nodes that we handle do that. This
%%   is not really a problem, because all of our commands have one
%%   output, which if they dont open this would cause problems in the
%%   sequential script too.}

@@

As mentioned in \Cref{parallelizability}, commands in \tpur can be
parallelized using a map stage and a \todo{fold} stage. We have
implemented these stages for several commands in \tpur to enable \sys
to parallelize them. Some interesting examples are \ttt{sort}, which
can be parallelized simply by using \ttt{sort} as a map and \ttt{sort
  -m} as \todo{fold}, and \todo{wc} \TODO{Niko, can you complete
  that with wc or any other interesting example?}
